Analysis-ready VCF at Biobank scale using Zarr

### 2024-11-14 paper.tex
Add spruce citation and update text

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a haploid genome assembly size
of 17.7 Gbp. Here, we assess the performance of \texttt{vcf2zarr} as
applied to 1,063 whole-genome resequenced spruce individuals from a
recent study (in preparation). The dataset consists of approximately
3,745 million single-nucleotide variants and small indels from the 12
autosomes (15.9 Gbp) totalling 7.4TiB of VCF data after \texttt{bgzip}
compression, distributed over 165 VCF files. For computational and
storage reasons, the genotypes only include the `PL` field.

%% Regarding data storage files, which to use?
%% find spruce.icf -name "*" | wc -l -> 1394080
%% # chunks in spruce.icf.inspect.txt -> 1342569
%% du -hs spruce.icf -> 6.8TiB
%% sum size in spruce.icf.inspect.txt -> 6.8TiB
%% du -hs spruce.vcz -> 
%% sum size in spruce.vcz.inspect.txt -> 6.6TiB
%% # chunks in spruce.vcz.inspect.txt -> 11984581
%% More chunk files despite smaller input data: could selection of
%% chunk size matter here?
Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.4TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 2,048 GB RAM and 128 cores. Conversion to ICF required
13h45min runtime. The ICF representation used a total of 6.8 TiB over
1,394,080 data storage files. Conversion to Zarr required 18h51min
runtime, generating a dataset with 32 arrays, consuming a total of XXX
TiB storage over 11,984,581 chunk files. The reduction in total storage (XX)
is negligible, presumably due to the small sample size. The top fields
in terms of storage are detailed in Table~\ref{tab-spruce-data}.

>>>>

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3,745 million
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. Since the dataset originally was
generated for downstream analyses that require genotype likelihoods,
the genotypes include the PL field.

%% Regarding data storage files, which to use?
%% find spruce.icf -name "*" | wc -l -> 1394080
%% # chunks in spruce.icf.inspect.txt -> 1342569
%% du -hs spruce.icf -> 6.8TiB
%% sum size in spruce.icf.inspect.txt -> 6.8TiB
%% du -hs spruce.vcz -> 6.7TiB
%% sum size in spruce.vcz.inspect.txt -> 6.6TiB
%% # chunks in spruce.vcz.inspect.txt -> 11984581
%% More chunk files despite smaller input data: could selection of
%% chunk size matter here?
Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.4TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 2,048 GB RAM and 128 cores. Conversion to ICF required
13h45min runtime. The ICF representation used a total of 6.8 TiB over
1,394,080 data storage files. Conversion to Zarr required 18h51min
runtime, generating a dataset with 32 arrays, consuming a total of 6.7
TiB storage over 11,984,581 chunk files. The reduction in total
storage (1.1X) is negligible, presumably due to the small sample size.
The top fields in terms of storage are detailed in
Table~\ref{tab-spruce-data}.
